A formal treatment of bidirectional typing

### 2023-10-13 tex/ESOP24/S1-intro.tex
[ tex ] Revised S1

Our work adopts a proof-relevant approach to (bidirectional) type synthesis, as illustrated by \citet{Wadler2022} for \PCF.
The proof-relevant formulation deviates from the usual one:
traditionally, a type synthesis algorithm is presented as \emph{algorithmic rules}, for example in the form $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, which denotes that annotations can be added to~$\isTerm{t}$ in the surface language%
\Josh{The algorithm adds annotations to the input term?}
to produce $\isTerm{t}'$ of type~$A$ in the core language.
Such an algorithm is then accompanied by soundness and completeness assertions that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised.
By contrast, the proof-relevant approach exploits the simultaneously computational and logical nature of Martin-L\"of type theory and formulates \emph{algorithmic soundness, completeness, and decidability in one go}.
We explain this in a bit more detail below.
%(although they do not emphasise the difference of their approach from the traditional one)

@@

%
In short, we develop a general and formal theory of bidirectional type synthesis for simply typed languages, including 
\begin{enumerate}
  \item general definitions for bidirectional type systems and mode-correctness;
  \item mode derivations for explicitly dealing with annotations in the theory (in particular leading to a more natural and useful version of completeness than annotatability~\cite{Dunfield2021}), and mode decoration for helping the user to work with annotations in practice;
  \item rigorously proven consequences of mode-correctness---the uniqueness of synthesised types and the decidability of bidirectional type synthesis, which amounts to
  \item a fully verified generator for proof-relevant type-synthesisers.
\end{enumerate}
Our theory was first developed in \Agda and then translated to the mathematical vernacular, filling the gap between our informal presentation and its verified implementation.

>>>>

Our work adopts a proof-relevant approach to (bidirectional) type synthesis, as illustrated by \citet{Wadler2022} for simply typed $\lambda$-calculus with natural numbers.
The proof-relevant formulation deviates from the usual one:
traditionally, a type synthesis algorithm is presented as \emph{algorithmic rules}, for example in the form $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, which denotes that annotations can be added to~$\isTerm{t}$ in the surface language%
\Josh{The algorithm adds annotations to the input term?}
to produce $\isTerm{t}'$ of type~$A$ in the core language.
Such an algorithm is then accompanied by soundness and completeness assertions that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised.
By contrast, the proof-relevant approach exploits the simultaneously computational and logical nature of Martin-L\"of type theory and formulates \emph{algorithmic soundness, completeness, and decidability in one go}.
We explain this in a bit more detail below.
%(although they do not emphasise the difference of their approach from the traditional one)

@@

%
In short, we develop a general and formal theory of bidirectional type synthesis for simply typed languages, including 
\begin{enumerate}
  \item general definitions for bidirectional type systems and mode-correctness;
  \item mode derivations for explicitly dealing with annotations in the theory (in particular leading to a more natural and useful version of completeness than \varcitet{Dunfield2021}{'s annotatability}), and mode decoration for helping the user to work with annotations in practice;
  \item rigorously proven consequences of mode-correctness---the uniqueness of synthesised types and the decidability of bidirectional type synthesis, which amounts to
  \item a fully verified generator for proof-relevant type-synthesisers.
\end{enumerate}
Our theory was first developed in \Agda and then translated to the mathematical vernacular, filling the gap between our informal presentation and its verified implementation.
